---
title: "Coursera Project"
format: 
  html:
    code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
---

## Instructions

Instructions on this project are given on [Coursera](https://www.coursera.org/learn/crash-course-in-causality/exam/MbcoE/data-analysis-project-analyze-data-in-r-using-propensity-score-matching/attempt).

Objective is to estimate the causal effect of a labor training program on post-intervention income levels.


## Project setup

Setup consists of installing the required R-packages `tableone`, `Matching` and 
`MatchIt` and loading the sample data `lalonde` which is included in the 
`MatchIt` package. 

```{r}
#| label: setup
#| message: false
pkg = c("tableone", "Matching", "MatchIt")
pkg = pkg[which(!pkg %in% available.packages()[, 1])]

install.packages(pkg)

library("data.table")
library("ggplot2")
library("tableone")
library("Matching")
library("MatchIt")
data(lalonde) 

data = as.data.table(lalonde)
head(data)
```


## Exercises

### Standardized mean difference

Calculate standardized differences for all confounding variables and obtain 
selected values.

```{r}
#| label: smd

smd = CreateTableOne(
  vars = setdiff(names(data), y = "treat")
  , strata = "treat"
  , data = data
  , test = FALSE
)

print(smd, smd = TRUE)
```

**Answers:**

The pre-matching standardized difference for `married` is 
`r round(attributes(smd$ContTable)$smd["married",], digits = 2L)`.  

The unadjusted difference in `real earnings in 1978` between treated and 
untreated is 
`r round(smd$ContTable[["1"]]["re78", "mean"] - smd$ContTable[["0"]]["re78", "mean"], digits = 0L)`$.


### Propensity score

Estimate the propensity score for each subject using logistic regression and 
return a summary of the distribution of the propensity scores.

```{r}
#| label: pscore

pmod = glm(
  treat ~ .
  , family = "binomial"
  , data = data[, .SD, .SDcols = setdiff(names(data), y = "re78")]
)

pscore = pmod$fitted.values
summary(pmod)
```

**Answers:** 

The range of the propensity scores is [`r round(range(pscore), digits = 3L)`].


### Propensity score matching

Match treated with control subjects based on the propensity score, match 
one-by-one, no replacement, no caliper. 

```{r}
#| label: pscore-matching

set.seed(931139)

pmatch = Match(
  Tr = data$treat     # treatment variable name
  , X = pscore        # propensity score
  , M = 1L            # number of matches
  , replace = FALSE   # no replacement
)

summary(pmatch)
```

Subset to matched data and re-calculate standardized mean differences. 

```{r}
#| label: smd-match

dmatch = data[c(pmatch$index.treated, pmatch$index.control)]

smd = CreateTableOne(
  vars = setdiff(names(dmatch), y = "treat")
  , strata = "treat"
  , data = dmatch
  , test = FALSE
)

print(smd, smd = TRUE)
```

**Answers:**

The post-matching standardized difference for `married` is 
`r round(attributes(smd$ContTable)$smd["married",], digits = 2L)`.  

The largest standardized difference is observed for race `black`.


### Propensity score matching with caliper

Match treated with control subjects based on the propensity score, match 
one-by-one, no replacement, **set caliper to 0.1**. 

```{r}
#| label: pscore-matching-caliper

set.seed(931139)

pmatch = Match(
  Tr = data$treat     # treatment variable name
  , X = pscore        # propensity score
  , M = 1L            # number of matches
  , replace = FALSE   # no replacement
  , caliper = 0.1     # caliper
)

summary(pmatch)
```

```{r}
#| label: smd-match-caliper

dmatch = data[c(pmatch$index.treated, pmatch$index.control)]

smd = CreateTableOne(
  vars = setdiff(names(dmatch), y = "treat")
  , strata = "treat"
  , data = dmatch
  , test = FALSE
)

print(smd, smd = TRUE)
```

**Answers:**

There are `r length(pmatch$index.treated)` matched pairs.


### Outcome analysis

```{r}
#| label: outcome-analysis

diffy = data$re78[pmatch$index.treated] - 
  data$re78[pmatch$index.control]

test = t.test(diffy)
test
```

**Answers:**

The adjusted difference in `real earnings in 1978` between treated and 
untreated is 
`r round(smd$ContTable[["1"]]["re78", "mean"] - smd$ContTable[["0"]]["re78", "mean"], digits = 0L)`$.

The 95% confidence interval of the differences in earnings is 
[`r round(test$conf.int, digits = 2L)`]$.
